Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-01-27 src/tex/ms.tex
[showyourwork] overleaf sync

   \author{Tobias Buck\inst{1,2}
          \and
          Berkay Günes\inst{1,2}
          }

@@

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Chemical abundances of stars are able to reveal important galactic parameters, such as the initial mass function high-mass slope and the frequency of type Ia supernovae. Constraining these parameters is of critical importance in order to create realistic hydrodynamical simulations and understand galactic physics. However, so far inference was significantly limited by the computational cost of traditional Bayesian algorithms. Here we present a new method using simulation-based inference (SBI) to circumvent previous limitations and enable inference from massive stellar surveys.
    SBI enables a new approach to the inference of the posterior distribution of Galactic parameters under an intractable likelihood function by utilizing forward simulations of stellar abundance distributions.
    Using SBI we are able to predict global galactic parameters of great accuracy and precision from chemical abundances of multiple stars, faster than conventional methods like Hamiltonian Monte Carlo inference. Our method easily scales to hundred-thousands of stars and beats previous approaches by orders of magnitude in compute time and accuracy.}
  % aims heading (mandatory)
   {}
  % methods heading (mandatory)
   {}
  % results heading (mandatory)
   {}
  % conclusions heading (optional), leave it empty if necessary 
   {}

@@

Currently all cosmological simulations of galaxy formation \citep[e.g.][]{Sawala2016,Hopkins2018,Pillepich2018,Buck2020,Buck2020c,Font2020,Agertz2021} rely on a small set of galactic parameters to describe fundamental effects of stellar evolution, including the birth and death rates for various types of stars or the production of heavy elements inside stars \citep[e.g.][]{Buck2021}. Two crucial unknowns are the shape of the initial mass function (IMF), setting the mass distribution of stars born from the interstellar medium (ISM), and the rate of Type Ia supernovae (SN\,Ia) explosions.

The exact values of parameters sensitively affect chemical evolution tracks \citep{2005A&A...430..491R,2015MNRAS.449.1327V,2015MNRAS.451.3693M}, yet their observational constraints remain weak.
%
For example, a range of high-mass IMF slopes have been suggested \citep[Tab.\,7]{2016ApJ...824...82C}, with a steeper-than-canonical slope being suggested by a range of studies \citep[e.g.][]{2015ApJ...806..198W,Rybizki2015,Chabrier2014}. In addition, the IMF slope may itself be not a constant but rather a function of metallicity, introducing further complexity \citep[e.g.][]{2019MNRAS.482..118G,Martin2019}. Similarly, the choice of SN\,Ia delay-time-distribution and normalization plays a crucial role in the enrichment of the ISM \citep[e.g.][]{Buck2021} and is heavily debated \citep{2010ApJ...722.1879M,2012MNRAS.426.3282M,2015ApJ...810..137J}.

The goal of this work is to demonstrate how we can use modern machine learning (ML) techniques in tandem with a simple galactic chemical evolution (GCE) model in a Bayesian framework to infer global galactic parameters from a set of stars. Our work focuses on two key parameters; the high-mass slope of the \citet[Tab.\,1]{2003PASP..115..763C} IMF and the rate of SN\,Ia explosions per unit mass. We assume that both are constant across the galaxy and time.

A methodologically very similar work was previously presented by \cite{Philcox_2019} using Hamilton Monte Carlo sampling for the inference. The result of this work was an impressive speed-up over classical Markov Chain Monte Carlo sampling with no loss of accuracy. Here we make use of simulation-based inference \citep[SBI, e.g.][]{Cranmer2020}. For high-dimensional posterior functions, SBI gives tremendous computational improvements over more traditional methods such as Markov Chain Monte Carlo (MCMC) or even the more efficient Hamilton Monte Carlo (HMC) methods.

This paper is structured as follows. In Section~\ref{sec:methods} we detail our methods. In particular, we discuss the GCE model and the methods behind SBI. Our results are presented in Section~\ref{sec: Results} where we test the performance of our approach on mock data and data taken from a hydrodynamical simulation. We conclude our study with Section~\ref{sec: conclusion} with a summary and outlook of our results.

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the Ilustris TNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this missmatch are examined in Sec.\,\ref{subsec: mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{1998A&A...334..505P} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

We train our NPE with $10^5$ data points consisting of $n_{\rm elemets}=8$ simulated with the NN emulator described in the previous section with inputs sampled from the priors shown in Tab. \ref{tab:priors}. Training takes $\lesssim10$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPUs. 
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimick observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_posterior_NPE_C.pdf}
     \vspace{-.5cm}
     \caption{Absolute percentage error of the neural posterior density estimate for a single star. Different coloured histograms show the full error distribution for all 6 parameters of interest with the median values highlighted by the vertical dashed lines. The box plots show again the first and third quartiles with the median represented by a vertical line and the whiskers extending from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The global parameters of main interest for this work are shown by the light blue and red histogram.}
     \label{fig:posterior_APE}
     \script{plot_posterior_APE.py}
\end{figure}

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $10$ seconds to build a posterior for all six parameters for a dataset size of 1000 stars (each time sampling the single star posterior with 1000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is hence 610 seconds for 1000 stars hence our methods making it more than $1.180$ times faster than current HMC methods \citep[cf.][who needs $40$h for only 200 stars]{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim72.000$. \textcolor{red}{check if those speed-up values are correctly calculated...}
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_TNG_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainities as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions).}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_tng_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_alternative_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainities as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions).}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_alt_yield_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/TNG_sim_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainities as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions).}
    \label{fig:TNG_N_star_analysis}
    \script{tng_inference.py}
\end{figure*}

\paragraph{Inference on mock data from \texttt{CHEMPY} with TNG yield set}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}

@@

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma levels$.

Using a sample of 1000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the correct value.

@@

\subsection{Inference on mock data from \texttt{CHEMPY} with incorrect yield set}
\label{subsec: mocks_wrong_yield}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_alternative_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_alt_yield_inference.py}
\end{figure}

\subsection{Inference on mock data from the IllustrisTNG simulation}
\label{subsec: tng_sim}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/TNG_simulation.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{tng_inference.py}
\end{figure}

\section{Discussion}
\label{sec: conclusion}

\textcolor{red}{here goes the assumption that stars are iid since each stars is produced by a single call to chempy. However, real physics might correlate stars abundacnes by star formation histories and cumulative enrichment...}

>>>>

   \author{Tobias Buck\inst{1,2}
          \and
          Berkay Günes\inst{1,2}
          \and
          et al.
          }

@@

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to the large datasets produced by modern stellar surveys.}
  % aims heading (mandatory)
   {This study aims to leverage simulation-based inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances. By focusing on the high-mass IMF slope and SN\,Ia normalization, we aim to demonstrate the advantages of SBI over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combined a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train an SBI pipeline. Mock datasets were generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, and additional tests were conducted on data from a Milky Way-like galaxy taken from a hydrodynamical simulation. SBI results were benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {Our SBI approach achieves a $\sim1,000\times$ ($\sim72,000\times$) speed-up compared to HMC, reducing inference runtime from $\sim40$ hours for 200 stars to mere minutes for thousands of stars. Inference on $1,000$ stars yielded precise estimates for the IMF slope ($\alpha_{\rm IMF} = −2.296 \pm 0.009$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = −2.889 \pm 0.010$), deviating by less than 0.04\% from the ground truth. SBI also demonstrated greater robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in galactic chemical enrichment studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys, such as 4MOST and WEAVE, facilitating deeper insights into the chemical and dynamical evolution of galaxies.}

%Chemical abundances of stars are able to reveal important galactic parameters, such as the initial mass function high-mass slope and the frequency of type Ia supernovae. Constraining these parameters is of critical importance in order to create realistic hydrodynamical simulations and understand galactic physics. However, so far inference was significantly limited by the computational cost of traditional Bayesian algorithms. Here we present a new method using simulation-based inference (SBI) to circumvent previous limitations and enable inference from massive stellar surveys.
%SBI enables a new approach to the inference of the posterior distribution of Galactic parameters under an intractable likelihood function by utilizing forward simulations of stellar abundance distributions.
%Using SBI we are able to predict global galactic parameters of great accuracy and precision from chemical abundances of multiple stars, faster than conventional methods like Hamiltonian Monte Carlo inference. Our method easily scales to hundred-thousands of stars and beats previous approaches by orders of magnitude in compute time and accuracy.

@@

Understanding the chemical enrichment of galaxies is fundamental to deciphering their formation and evolution. Chemical abundances of stars offer a wealth of information about galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. These parameters critically influence the production of heavy elements \citep[e.g.][]{2005A&A...430..491R,2015MNRAS.449.1327V,2015MNRAS.451.3693M}, stellar feedback, and star formation histories, making their accurate determination essential for realistic hydrodynamical simulations of galaxy formation \citep[e.g.][]{Sawala2016,Hopkins2018,Pillepich2018,Buck2020,Buck2020c,Buck2021,Font2020,Agertz2021}. Despite their importance, constraining these parameters has proven challenging due to limited observational data and the computational demands of traditional inference techniques.

For example, a range of high-mass IMF slopes have been suggested \citep[Tab.\,7]{2016ApJ...824...82C}, with a steeper-than-canonical slope being suggested by a range of studies \citep[e.g.][]{2015ApJ...806..198W,Rybizki2015,Chabrier2014}. In addition, the IMF slope may itself be not a constant but rather a function of metallicity, introducing further complexity \citep[e.g.][]{2019MNRAS.482..118G,Martin2019}. Similarly, the choice of SN\,Ia delay-time-distribution and normalization plays a crucial role in the enrichment of the ISM \citep[e.g.][]{Buck2021} and is heavily debated \citep{2010ApJ...722.1879M,2012MNRAS.426.3282M,2015ApJ...810..137J}.

Recent advances in stellar spectroscopic surveys, such as APOGEE and GALAH, have produced unprecedented datasets of stellar chemical abundances. These datasets hold the potential to unlock detailed constraints on galactic parameters across diverse environments. However, traditional Bayesian inference methods, such as Markov Chain Monte Carlo (MCMC) and Hamiltonian Monte Carlo (HMC), struggle to scale to these large datasets. Such methods are computationally expensive, requiring hours of runtime for even modest sample sizes, and are susceptible to biases when confronted with high-dimensional posterior distributions.

In this work, we present a novel approach leveraging simulation-based inference \citep[SBI, e.g.][]{Cranmer2020} to address these limitations. SBI bypasses the need for explicit likelihood functions, enabling efficient and scalable inference of galactic parameters directly from simulated stellar abundances. By combining a neural network emulator for the \texttt{CHEMPY} Galactic Chemical Evolution (GCE) model with a Neural Posterior Estimator (NPE), we achieve rapid and robust inference. Unlike HMC, which requires extensive sampling for each dataset, our method amortizes the computational cost during training, allowing subsequent inference to scale seamlessly to larger datasets.

This study focuses on two critical global galactic parameters: 
the high-mass slope of the \citet[Tab.\,1]{2003PASP..115..763C} IMF ($\alpha_{\rm IMF}$) and the SN\,Ia normalization ($\log_{10}(N_{\rm Ia})$), the rate of SN\,Ia explosions per unit mass. We demonstrate the accuracy, scalability, and robustness of SBI through tests on mock datasets generated by \texttt{CHEMPY} \citep{Rybizki_2017}, as well as on data from hydrodynamical simulations. Additionally, we compare our results to those obtained using HMC-based inference on the same datasets \citep[see][]{Philcox_2019}, highlighting SBI's superior performance in terms of speed, precision, and resilience to model misspecification.

The structure of this paper is as follows: In Section~\ref{sec:methods}, we outline the methods used, including the GCE model and SBI framework. Section~\ref{sec: Results} presents our results on both \texttt{CHEMPY} and IllustrisTNG \citep{Pillepich2018} mock data, emphasizing SBI's advantages over traditional approaches. Finally, in Section~\ref{sec: discussion}, we discuss the broader implications of our findings and outline potential future applications, before concluding in Section~\ref{sec: conclusion}.

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IlustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

We train our NPE with $10^5$ data points consisting of $n_{\rm elemets}=8$ simulated with the NN emulator described in the previous section with inputs sampled from the priors shown in Tab. \ref{tab:priors}. Training takes $\lesssim10$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPUs. 
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_posterior_NPE_C.pdf}
     \vspace{-.5cm}
     \caption{Absolute percentage error of the neural posterior density estimate for a single star. Different colored histograms show the full error distribution for all 6 parameters of interest with the median values highlighted by the vertical dashed lines. The box plots show again the first and third quantiles with the median represented by a vertical line and the whiskers extending from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The global parameters of main interest for this work are shown by the light blue and red histogram.}
     \label{fig:posterior_APE}
     \script{plot_posterior_APE.py}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_TNG_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainties as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions). See Sec.~\ref{subsec:chempy_tng}.}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_tng_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_alternative_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data created with a different yield set than the training data. See Sec.~\ref{subsec:mocks_wrong_yield}.}
    \label{fig:CHEMPY_alt_N_star_analysis}
    \script{chempy_alt_yield_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/TNG_sim_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data taken from an IlustrisTNG Milky Way-like galaxy. See Sec.~\ref{subsec:tng_sim}.}
    \label{fig:TNG_N_star_analysis}
    \script{tng_inference.py}
\end{figure*}

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $10$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is hence 610 seconds for 1,000 stars hence our methods making it more than $1.180$ times faster than current HMC methods \citep[cf.][who needs $40$h for only 200 stars]{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim72,000$. \textcolor{red}{check if those speed-up values are correctly calculated...}
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

@@

\subsection{Inference on mock data from \texttt{CHEMPY} with TNG yield set}
\label{subsec:chempy_tng}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1,000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}

@@

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma$ levels.
%
Using a sample of 1,000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the correct value.

@@

\subsection{Inference on mock \texttt{CHEMPY} data with incorrect yield set}
\label{subsec:mocks_wrong_yield}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_alternative_yields.png}
    \vspace{-.25cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_sbi} but for the mock data created with a different yield set than the training data. See Sec.~\ref{subsec:mocks_wrong_yield}.}
    \label{fig:CHEMPY_alt_sbi} 
    \script{chempy_alt_yield_inference.py}
\end{figure}

There is quite some discussion in the literature about stellar nucleosynthesis with various different yield sets proposed \citep[see e.g. discussion in][]{Rybizki_2017}. In fact, any tabulated yield set will differ from reality and during an application of our inference pipeline it will not be clear which tabulated yield set most closely matches reality and should be used.
In order to investigate how sensitive our method is to a model missspecification by using an incorrect yield set, we create another set of mock data using \texttt{CHEMPY} with a different yield set than during training of our NPE. For better cross-comparison we decided to use the same alternative yield set as presented in Tab.~5 of \citep{Philcox_2019}. In short this yield set uses SN Ia yields from \citet{2003NuPhA.718..139T}, SN II yields from \citet{2013ARA&A..51..457N} and AGB yields from \citet{2016ApJ...825...26K}. By chosing this set of yields we have made sure that contributions to all three processes differ by $\mathcal{O}(10\%)$ \citep[see Sec.~6.2 of][for more details]{Philcox_2019}.

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/TNG_simulation.png}
    \vspace{-.25cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_sim_sbi} but for the mock data taken from an IlustrisTNG Milky Way-like galaxy. See Sec.~\ref{subsec:tng_sim}.}
    \label{fig:CHEMPY_TNG_sim_sbi} 
    \script{tng_inference.py}
\end{figure}

Our mock data generation and inference then follows the one of Sec.~\ref{subsec:chempy_tng}. Note, this means we apply our NPE trained on data with the TNG yield set to data created with the above mentioned alternative yield set. This effectively probes the effect of model missspecification on the inference results.

We show our inference results for this setup for varying number of stars (blue) in Fig.~\ref{fig:CHEMPY_alt_N_star_analysis} comparing again against HMC results (red) from \citet{Philcox_2019}. Note that the SBI results are less biased than the HMC results especially for $\alpha_{\rm IMF}$. In fact with 200 stars HMC inferences are about \textcolor{red}{calculate value} $\sigma$ away from the ground truth value for $\alpha_{\rm IMF}$ and $\log_{10}(N_{\rm Ia})$, respectively. Again, the joint posterior is shown in Fig.~\ref{fig:CHEMPY_alt_sbi} and shows that while individually parameter inferences look good, jointly taken the SBI results are on the edge of being $3\sigma$ away from the ground truth.

Nevertheless, the performance of our inference is still very good. We see that an increasing number of observations helps to decrease the models uncertainty just as before. However, we also note that our inference is now slightly biased as the observational data does not match the training data. Looking at Fig.~\ref{fig:CHEMPY_alt_sbi} we see that our inference is still consistent with the ground truth within the $3\sigma$ levels. Comparing this to the HMC results, we see that SBI is performing much better when the model is misspecified and leads to significantly less biased results. This is next to drastically reduced compute times another a key advantage of our SBI method compared to more standard approaches such as HMC.

In an accompanying paper we more closely investigate measures of model misspecification and inference of best fitting models next to just parameter inference.

\subsection{Inference on mock data from the IllustrisTNG simulation}
\label{subsec:tng_sim}

As a galactic chemical evolution code, \texttt{CHEMPY} is a one-zone model with simplified ISM physics that only approximately describes star formation and feedback as well as metal mixing in the ISM.
In the parametrization of \texttt{CHEMPY} as used here, we can assign each star to its own ISM environment, but we cannot exchange gas between environments and do not model sudden star formation or infall events. Hence, this section is dedicated to investigate whether this significantly biases our inference of the SSP parameters (noting that results from \citet{2019ApJ...874..102W} justify the treatment of ISM parameters as latent variables).

In order to explore what effect this simplified treatment of ISM physics has on the inference we now turn to a more complex model of the formation and chemical enrichment of a Milky Way-type galaxy taken from the IllustrisTNG simulations. Note, that by now also the NIHAO simulations \citep{Wang2015, Buck2020,Buck2020c} have implemented \texttt{CHEMPY} supported yield tables including the TNG yield set \citep{Buck2021} and hence would make for a nice dataset for our analysis. However, we have decided to use the exact same galaxy as in \citet{Philcox_2019} for better comparison of our results.

In detail, we use a single galaxy from the $z = 0$ snapshot of the highest-resolution TNG100-1 simulation. We choose a subhalo (index 523071) with mass close to $10^{12}\,\mathrm{M}_\odot$ to select a Milky Way-like galaxy. From this, we extract a set of 1,000 random `stellar particles' from a total of $\sim 40,000$. Each star particle has a mass of $\sim 1.4\times 10^6 \mathrm{M}_\odot$ \citep{2019ComAC...6....2N}. These act as proxies for stellar environments, giving the elemental mass fractions, $\{d_i^j\}$, and cosmological scale factor, $a_i$, at the time of stellar birth. Mass fractions are converted to [X/Fe] abundance ratios using \citet{2009ARA&A..47..481A} solar abundances as in \textit{Chempy}, with the scale-factor ($a_i$) to birth-time ($T_i$) conversion performed using \texttt{astropy} \citep{astropy:2013,astropy:2018},\footnote{\href{http://www.astropy.org}{http://www.astropy.org}} assuming a $\Lambda$CDM cosmology with \citet{planck2015} parameters, as in TNG \citep{2018MNRAS.475..648P}.\footnote{Note, as for the \textit{Chempy} mock data, we exclude any particles with $T_i\notin[2,12.8]\,\mathrm{Gyr}$ to ensure that the true times are well separated from our training age limits, avoiding neural network errors. This removes $\sim5\%$ of the stars.} Observational errors are incorporated as above, giving a full data-set that is identical in structure to the \texttt{Chempy} mock data. For more details and a plot of the [Mg/Fe] vs. [Fe/H] plane for this galaxy see Sec.~6.3 and Fig.~5 of \citet{Philcox_2019}.

We note that this TNG galaxy was deliberately chosen by \citet{Philcox_2019} to have both a high-$\alpha$ and low-$\alpha$ chemical evolution sequence to test their inference on a mock galaxy with Milky Way-like properties. There is still some debate on the exact formation of this bimodality but it is generally attributed to gas-rich mergers and different modes of star formation \citep[e.g.][]{2018MNRAS.474.3629G,2018MNRAS.477.5072M,2019MNRAS.484.3476C,Buck2020,Buck2023}. Similarly, in chemo-dynamical models, Milky Way-like bimodalities can also be achieved by a combination of radial migration and selection effects without the need for mergers or starbursts \citep[e.g.][]{2009MNRAS.396..203S,2013A&A...558A...9M,2017ApJ...835..224A}.

We show our inference results for the TNG data set in Fig.~\ref{fig:TNG_N_star_analysis} and Fig.~\ref{fig:CHEMPY_TNG_sim_sbi}, respectively.
Again, we find that inference becomes better with increasing number of stars. Despite the drastic difference in chemical enrichment model between training and testing data, our SBI pipeline is impressively capable of inferring the correct posterior values. SBI results for $\alpha_{\rm IMF}$ are almost perfect while $\log_{10}(N_{\rm Ia})$ is slightly biased low. This is exactly opposite to the results of HMC where inference for $\alpha_{\rm IMF}$ is biased and results for $\log_{10}(N_{\rm Ia})$ are more in agreement. Hence, our SBI inference is on par with the HMC results and given their extreme computational advantage they actually supersede HMC.

Looking at the joint posterior in Fig.~\ref{fig:CHEMPY_TNG_sbi} we see that also in this case SBI recovers the true parameters with less than $3\sigma$ bias.

\section{Discussion}
\label{sec: discussion}

\textcolor{red}{here goes the assumption that stars are iid since each star is produced by a single call to chempy. However, real physics might correlate stars abundances by star formation histories and cumulative enrichment...}

Our study demonstrates that simulation-based inference (SBI) provides a powerful and efficient alternative to conventional methods such as Hamiltonian Monte Carlo (HMC) for inferring global galactic parameters. By leveraging neural density estimators and neural network emulators, we achieved remarkable computational efficiency without compromising precision or accuracy.
